---
title: "Exploratory Analysis of OkCupid dataset"
author: "Winston Saunders"
date: "August 27, 2016"
output: 
    html_document:
        theme: united
        keep_md: true
---


##EXEC SUMMARY  
This explores several relationships in the OkCupid data [published on CRAN](https://cran.rstudio.com/web/packages/okcupiddata/index.html). Results explored include:  
0. Ages of OkCupid users.   
1. The Correlation of Religion and Drinking Habits.  
2. Changing drinking habits with age.  
3. Religious Affiliation with Age.  
4. TBD 

##Getting the Data
The OkCupid data is published on [CRAN](https://cran.rstudio.com/web/packages/okcupiddata/index.html) as a package for R suers. The data set consists of user profile data for 59,946 San Francisco OkCupid users (a free online dating website) from June 2012. The data are describded in the paper: Albert Y. Kim, Adriana Escobedo-Land (2015). OkCupid Profile Data for Introductory Statistics and Data Science Courses. Journal of Statistics Education, 23(2), which is found [here]( http://www.amstat.org/publications/jse/v23n2/kim.pdf)

```{r, echo=FALSE, message = FALSE}

## OKCupid Data Explore

## Winston Saunders



library(dplyr)
library(ggplot2)
library(RColorBrewer)
library(gridExtra)

```

The raw data is loaded as a library

```{r}
## load data
library(okcupiddata)
```

and it consist of these data fields (detailed descriptions of which can be found in the reference above). 

```{r, echo = TRUE}
## column names
profiles %>% colnames
```



## Basic Age and Sex Distributions 

The simplest and most obvious thing is to look at first is the sex and age distribution of OkCupid users in a histogram.

```{r echo=TRUE, fig.align='center', fig.height=3,fig.width=7.5}
    ## eliminate NAs and restrict age
    cleaned <- filter(profiles, !is.na(age), !is.na(sex), age > 18, age < 80) %>% 
            as_data_frame %>%
            select(sex, age)
    ## make data descriptive
    cleaned$sex<- cleaned$sex %>% gsub("m", "male", .) %>% gsub("f", "female", .)
   
    
```




```{r echo=FALSE, fig.align='center', fig.height=3,fig.width=7.5}

        ## plot age histogram
        age_hist <- ggplot(cleaned, aes(x= age, fill = sex)) + geom_bar(position = "dodge", stat = "count")  +
        ggtitle("okcupid: age distribution") +
        scale_fill_manual(values = c("#DD9834", "#3498DD")) + 
        ylab("count") + 
        xlab("age") +
        theme(legend.position="none")




        ## plot sex histogram
        sex_hist <- ggplot(cleaned, aes(x= sex, fill = sex)) + geom_bar(position = "dodge", stat = "count")  +
        ggtitle("okcupid: sex distribution") +
        scale_fill_manual(values = c("#DD9834", "#3498DD")) + 
        ylab("count") + 
        xlab("age")

        grid.arrange(sex_hist, age_hist, ncol=2)

```

```{r echo=FALSE, fig.align='center', fig.height=3,fig.width=7.5}

    ## compute age groups using mutate
    analyzed <- cleaned %>% mutate(age_group = 2 + 4 *  age %/% 4)
    analyzed$age_group <- analyzed$age_group %>% as.factor
    
    ## group_by data and summarize by sex and age
    sex_count <- group_by(analyzed[,c("age_group", "sex")], age_group, sex) %>% summarize(n_sex = n())
    ## count the total number of males and females
    age_count <- group_by(analyzed[,c("age_group")], age_group) %>% summarize(n_age = n())
    ## join the data
    analyzed <- left_join(sex_count, age_count, by = "age_group") %>% mutate(freq = n_sex/n_age, freq = ifelse(is.na(freq), 0, freq), delta_percent = 200*(freq - 0.5))

    

    ## compute the overall number of men and women
    n_males <- analyzed %>% filter(sex == "male")  
    n_males <- n_males$n_sex %>% sum()
    
    n_females <- analyzed %>% filter(sex == "female") 
    n_females <- n_females$n_sex %>% sum()

```

Overall the number of men in `r n_males` and the number of females is `r n_females`, with an overall ratio of about `r round(n_males/n_females, 1)`:1. 

A couple of notable features are that the distributions peaks in the late 20's and there is along tail on the distribution where the populations of men and women seem to be equal. 
### Digging deeper into the Male - Female Population

We can dig deeper into the differences in the age group distributions by looking at the differences between them as bucketed by age.

```{r echo=TRUE, fig.align='center', fig.height=3,fig.width=7.5}

    ## compute age_group factor using mutate
    analyzed <- cleaned %>% mutate(age_group = 2 + 4 *  age %/% 4)
    analyzed$age_group <- analyzed$age_group %>% as.factor
    
    ## group_by data and summarize by sex and age
    sex_count <- group_by(analyzed[,c("age_group", "sex")], age_group, sex) %>% summarize(n_sex = n())
    ## count the total number of males and females
    age_count <- group_by(analyzed[,c("age_group")], age_group) %>% summarize(n_age = n())
    ## join the data
    analyzed <- left_join(sex_count, age_count, by = "age_group") %>% mutate(freq = n_sex/n_age, freq = ifelse(is.na(freq), 0, freq), delta_percent = 200*(freq - 0.5))
```

A graph below shows clearly the diffences in the numbers of male and female users, with men outnumbering women by an approximately 25% for ages below 50, while in the 60's to 70's, women outnumber men by about 10%.  


```{r, echo = FALSE, fig.align='center', fig.height=3, warning = FALSE}
    
    
    analyzed <- analyzed %>% filter(sex == "male") %>% as.data.frame
    
    
    n.color <- length(unique(analyzed$age_group))
    getPalette = colorRampPalette(brewer.pal(8, "Dark2"))

    p <- ggplot(analyzed, aes(x = age_group, y = delta_percent, group = sex, fill = age_group)) +
        geom_bar(stat="identity", alpha = 1.0) +
        ylab("percent difference") +
        scale_fill_manual(values = getPalette(n.color)) +
        xlab("age") +
        ggtitle("(male - female) population difference by age") +
        theme(legend.position="none") 
    
print(p)
     
```


###Age distribution of male and female populations

The above data show differences male to female users by age, but they do not clearly reveal what the differences between the popluations are. We can explore more intrinsic male and female behavior by separating the male and females populations and normalizing them to the relevant number of users of each sex.  

```{r echo=FALSE, fig.align='center', fig.height=3,fig.width=7.5}

    ## compute age groups using mutate
    analyzed <- cleaned %>% mutate(age_group = 2 + 4 * floor(age/4))
    analyzed$age_group <- analyzed$age_group %>% as.factor
    
    ## group_by data and summarize by sex and age
    age_count <- group_by(analyzed[,c("age_group", "sex")], age_group, sex) %>% summarize(n_age = n())
    ## count the total number of males and females
    sex_count <- group_by(analyzed[,c("sex")], sex) %>% summarize(n_sex = n())
    ## join the data
    analyzed <- left_join(age_count, sex_count, by = "sex") %>% mutate(freq = n_age/n_sex, freq = ifelse(is.na(freq), 0, freq))
    
    ## data snapshot
    
    #    age_group    sex n_age n_sex        freq
    #       <fctr>  <chr> <int> <int>       <dbl>
    # 1         18 female   281 23955 0.011730328
    # 2         18   male   330 35680 0.009248879
    # 3         22 female  2838 23955 0.118472135
    # 4         22   male  3923 35680 0.109949552
    # 5         26 female  5325 23955 0.222291797
    
```



```{r echo=FALSE, fig.align='center', fig.height=3,fig.width=7.5}
    
    ## Generate Distribution and Cummulative Plots

    ## distribution plot is straight forward
    p_dist <- ggplot(analyzed, aes(x = age_group, y = 100*freq, fill = sex, group = sex)) +
        geom_bar(stat = "identity", position = "dodge") +
        #geom_bar(aes(fill = sex), position = "dodge") +
        ggtitle("okcupid: age distribution of sexes") +
        scale_fill_manual(values = c("#DD9834", "#3498DD")) + 
        ylab("percent of total per sex") + 
        xlab("age_group")
    
    
    ## Cummulative requires some manipuations
    ## first make a "wide" data set
    male <- analyzed %>% filter(sex == "male") %>% select(age_group, sex, freq)
    female <- analyzed %>% filter(sex == "female") %>% select(age_group, sex, freq)

    analyzed_wide <- inner_join(male, female, by = "age_group") %>% mutate(pop_diff = freq.x - freq.y)

    ## compute cummulative sums

    analyzed_wide2 <- analyzed_wide %>% as.data.frame %>% mutate(male_cummul = cumsum(freq.x), female_cummul = cumsum(freq.y))

    ## plot the data
    p_cummul <- ggplot(analyzed_wide2, aes(x = age_group, y = 100*male_cummul, group = sex.x)) +
        geom_line(size = 1.5, color = "#3498DD") +
        geom_line(aes(y = 100*female_cummul, group = sex.y), size = 1.5, color = "#DD9834") +
        ylab("cummulative percent") +
        xlab("age") +
        ggtitle("cummulative populations by age") 
    


```


Both populations have a gamma distribution-like shape with a mean near 28, but have some subtle differences as well.



```{r echo=FALSE, fig.align='center', fig.height=3,fig.width=7.5}
grid.arrange(p_dist, p_cummul, ncol=2)
```

For instance, about 50% of the of the users of both sexes are between the ages of 24 and 32 (recall the bins in this case are four years wide), and between the ages of 32 and 50 the relative frequency of both men and women are the same. For women, those above age 50 women represent a higher proportion of the female population than men do for the male population.

The cummulative plot highlights the longer tail on the female age distribution. <br>
A normalized bar chart shows the differences in age population by sex, again with women being more numerous at ages above about 50. 

```{r echo=FALSE, fig.align='center', fig.height=3,fig.width=7.5}

## GENERATE PLOT

    n.color <- length(unique(analyzed$age_group))
    getPalette = colorRampPalette(brewer.pal(8, "Dark2"))
    

    p <- ggplot(analyzed, aes(x = sex, y = 100*freq, fill = age_group)) +
        geom_bar(stat="identity") +
        ggtitle("okcupid: age and sex") + 
        scale_fill_manual(values = getPalette(n.color)) + 
        ylab("percent") + 
        xlab("") +
        coord_flip()+
        theme(legend.position="bottom") +
        guides(fill=guide_legend(nrow=2))
        
    
    print(p)
    
    
```

## Religion  and Drinking Habits

The profile data contains information the religious affiliation of users as well as their drinking habits. Let's see if there are relationships...  

#### Affiliation
The religious data contain various statements of both 'affiliation' and what I will call the 'devoutness' of that affilation. 

```{r}
head(unique(profiles$religion), 10)
```


For the first analysis, I strip off the devoutness descriptors to focus on affilation (so, for instance, whether someone typed _"catholicism and somewhat serious about it"_, or _"catholicism and very serious about it"_, they would be regarded as having affiliation with catholicism).   
Drinking habits are characterized by self-selected ratings of _"not at all"_, ... _"socially"_, ..._"desperately"_.  
The data are cleaned by filtering NA's and then grouped and counted as above.

```{r echo=2:5, fig.align='center', fig.height=5,fig.width=7.5, message = FALSE}
    ## clean religion data into broad category
    cleaned <- filter(profiles, !is.na(drinks), !is.na(religion), !is.na(sex)) %>% as_data_frame
    
    ## get affiliation (strip devoutness modifiers) using gsub and simple regex
    cleaned$religious_affil <- gsub(' [A-z ]*', '', cleaned$religion) %>% as.factor()
    
    ## convert bnG$drinks to factor
    cleaned$drinks <- cleaned$drinks %>% as.factor()
    
    ## group the data 
    drinks_and_religion <- group_by(cleaned[,c("drinks", "religious_affil", "sex")], drinks, religious_affil, sex) %>% summarize(n = n())
    
    # Source: local data frame [106 x 4]
    # Groups: drinks, religious_affil [?]
    # 
    #         drinks religious_affil   sex     n
    #         <fctr>          <fctr> <chr> <int>
    # 1  desperately     agnosticism     f    25
    # 2  desperately     agnosticism     m    37
    # 3  desperately         atheism     f    18
    # 4  desperately         atheism     m    40
    # 5  desperately        buddhism     f     5
    # 6  desperately        buddhism     m    15
    # 7  desperately     catholicism     f     4
    # 8  desperately     catholicism     m     7
    # 9  desperately    christianity     f     7
    # 10 desperately    christianity     m     7
    # # ... with 96 more rows
    
    
    religion_count <- group_by(cleaned[,c("religious_affil", "sex")], religious_affil, sex) %>% summarize(group_count = n())
    

    
    analyzed <- left_join(drinks_and_religion, religion_count) %>% mutate(freq = n/group_count, freq = ifelse(is.na(freq), 0, freq))

 
    ## Clean up for display
    
    ## reorder factor
    analyzed$drinks <- factor(analyzed$drinks, levels = c("not at all", 
                                                           "rarely",
                                                           "socially",
                                                           "often",
                                                           "very often",
                                                           "desperately"))
    
    
    
    ## express sex in actual words
    
    analyzed$sex<- analyzed$sex %>% gsub("m", "male", .) %>% gsub("f", "female", .)

    ## make color palette
    
    n.color <- length(unique(analyzed$drinks))
    getPalette = colorRampPalette(brewer.pal(8, "Dark2"))
    
    
    ## plot
    p <- ggplot(analyzed, aes(religious_affil, weight = 100*freq)) +
        geom_bar(aes(fill = drinks), alpha = 0.9) +
        ggtitle("okcupid: drinking and religion") + facet_grid(sex~.) +
        coord_flip() + 
        scale_fill_manual(values = getPalette(n.color)) + 
        ylab("percent") + 
        xlab("religious affiliation")
    
    print(p)
```

Generally patterns are fairly similar between men and women. 

#### Highly and Non-Devout
Once the above machinery is in place, we can look for differences in the habits of the religiously devout and non-devout. To do this we just select for _"serious about it"_ and _"very serious about it"_ for the devout.


```{r echo=4, fig.align='center', fig.height=5,fig.width=7.5, message = FALSE}
    ## clean religion data into broad category
    cleaned <- filter(profiles, !is.na(drinks), !is.na(religion), !is.na(sex)) %>% as_data_frame

    cleaned <- filter(cleaned, grepl("somewhat serious", religion) | grepl("very serious", religion))
    
    

    ## get affiliation (strip modifiers) using gsub and simple regex
    cleaned$religious_affil <- gsub(' [A-z ]*', '', cleaned$religion) %>% as.factor()
    
    ## convert bnG$drinks to factor
    cleaned$drinks <- cleaned$drinks %>% as.factor()
    
    ## group the data 
    drinks_and_religion <- group_by(cleaned[,c("drinks", "religious_affil", "sex")], drinks, religious_affil, sex) %>% summarize(n = n())
    
    # Source: local data frame [106 x 4]
    # Groups: drinks, religious_affil [?]
    # 
    #         drinks religious_affil   sex     n
    #         <fctr>          <fctr> <chr> <int>
    # 1  desperately     agnosticism     f    25
    # 2  desperately     agnosticism     m    37
    # 3  desperately         atheism     f    18
    # 4  desperately         atheism     m    40
    # 5  desperately        buddhism     f     5
    # 6  desperately        buddhism     m    15
    # 7  desperately     catholicism     f     4
    # 8  desperately     catholicism     m     7
    # 9  desperately    christianity     f     7
    # 10 desperately    christianity     m     7
    # # ... with 96 more rows
    
    
    religion_count <- group_by(cleaned[,c("religious_affil", "sex")], religious_affil, sex) %>% summarize(group_count = n())
    

    
    analyzed <- left_join(drinks_and_religion, religion_count) %>% mutate(freq = n/group_count, freq = ifelse(is.na(freq), 0, freq))

 
    ## Clean up for display
    
    ## reorder factor
    analyzed$drinks <- factor(analyzed$drinks, levels = c("not at all", 
                                                           "rarely",
                                                           "socially",
                                                           "often",
                                                           "very often",
                                                           "desperately"))
    
    
    
    ## express sex in actual words
    
    analyzed$sex<- analyzed$sex %>% gsub("m", "male", .) %>% gsub("f", "female", .)

    ## make color palette
    
    n.color <- length(unique(analyzed$drinks))
    getPalette = colorRampPalette(brewer.pal(8, "Dark2"))
    
    
    ## plot
    p_devout <- ggplot(analyzed, aes(religious_affil, weight = 100*freq)) +
        geom_bar(aes(fill = drinks), alpha = 0.9) +
        ggtitle("devout") + facet_grid(sex~.) +
        coord_flip() + 
        scale_fill_manual(values = getPalette(n.color)) + 
        ylab("percent") + 
        xlab("religious affiliation")
    

```

For non-devout we select for _"not too serious"_ and _"laughing about it"_ prior to other cleaning 


```{r echo=4, fig.align='center', fig.height=5,fig.width=7.5, message = FALSE}
    ## clean religion data into broad category
    cleaned <- filter(profiles, !is.na(drinks), !is.na(religion), !is.na(sex)) %>% as_data_frame

    cleaned <- filter(cleaned, grepl("not too serious", religion) | grepl("laughing about it", religion))
    
    # # A tibble: 38,729 x 22
    #       age     body_type                diet
    #       <int>       <chr>               <chr>
    # 1     22 a little extra   strictly anything
    # 2     35        average        mostly other
    # 3     29        average     mostly anything
    # 4     31        average     mostly anything
    # 5     24           <NA>   strictly anything
    # 6     37       athletic     mostly anything
    # 7     28        average     mostly anything
    # 8     24           <NA>              <NA>
    # 9     30         skinny     mostly anything
    # 10    29           thin     mostly anything
    # # ... with 38,719 more rows, and 19 more
    # #   variables: drinks <chr>, drugs <chr>,
    # #   education <chr>, ethnicity <chr>,
    # #   height <int>, income <int>, job <chr>,
    # #   last_online <time>, location <chr>,
    # #   offspring <chr>, orientation <chr>,
    # #   pets <chr>, religion <chr>, sex <chr>,
    # #   sign <chr>, smokes <chr>, speaks <chr>,
    # #   status <chr>, essay0 <chr>
    

    ## get affiliation (strip modifiers) using gsub and simple regex
    cleaned$religious_affil <- gsub(' [A-z ]*', '', cleaned$religion) %>% as.factor()
    
    ## convert bnG$drinks to factor
    cleaned$drinks <- cleaned$drinks %>% as.factor()
    
    ## group the data 
    drinks_and_religion <- group_by(cleaned[,c("drinks", "religious_affil", "sex")], drinks, religious_affil, sex) %>% summarize(n = n())
    
    # Source: local data frame [106 x 4]
    # Groups: drinks, religious_affil [?]
    # 
    #         drinks religious_affil   sex     n
    #         <fctr>          <fctr> <chr> <int>
    # 1  desperately     agnosticism     f    25
    # 2  desperately     agnosticism     m    37
    # 3  desperately         atheism     f    18
    # 4  desperately         atheism     m    40
    # 5  desperately        buddhism     f     5
    # 6  desperately        buddhism     m    15
    # 7  desperately     catholicism     f     4
    # 8  desperately     catholicism     m     7
    # 9  desperately    christianity     f     7
    # 10 desperately    christianity     m     7
    # # ... with 96 more rows
    
    
    religion_count <- group_by(cleaned[,c("religious_affil", "sex")], religious_affil, sex) %>% summarize(group_count = n())
    

    
    analyzed <- left_join(drinks_and_religion, religion_count) %>% mutate(freq = n/group_count, freq = ifelse(is.na(freq), 0, freq))

 
    ## Clean up for display
    
    ## reorder factor
    analyzed$drinks <- factor(analyzed$drinks, levels = c("not at all", 
                                                           "rarely",
                                                           "socially",
                                                           "often",
                                                           "very often",
                                                           "desperately"))
    
    
    
    ## express sex in actual words
    
    analyzed$sex<- analyzed$sex %>% gsub("m", "male", .) %>% gsub("f", "female", .)

    ## make color palette
    
    n.color <- length(unique(analyzed$drinks))
    getPalette = colorRampPalette(brewer.pal(8, "Dark2"))
    
    
    ## plot
    p_non_devout <- ggplot(analyzed, aes(religious_affil, weight = 100*freq)) +
        geom_bar(aes(fill = drinks), alpha = 0.9) +
        ggtitle("non-devout") + facet_grid(sex~.) +
        coord_flip() + 
        scale_fill_manual(values = getPalette(n.color)) + 
        ylab("percent") + 
        xlab("religious affiliation")
    
    
```

```{r, fig.align='center', fig.height=5, fig.width=9, echo=FALSE  }

grid.arrange(p_devout, p_non_devout, ncol = 2)

```


An overall decrease in social drinking is observed, but in some cases this is due to an increase in both infrequent and frequent drinking. 

##Age and Drinking


The manchinery above is also easily adapted to exploring the relationship of drinking and age. In this case a very clear pattern emerges for both men and women, with a pronounced  tendency toward lighter drinking with increased age. 

```{r echo=4, fig.align='center', fig.height=5,fig.width=7.5, message = FALSE}
    
    ## clean data into broad category
    cleaned <- filter(profiles, !is.na(drinks), !is.na(age), !is.na(sex)) %>% as_data_frame
    
    ## set factor order
    cleaned$drinks <- factor(cleaned$drinks, levels = c("not at all", 
                                                            "rarely",
                                                            "socially",
                                                            "often",
                                                            "very often",
                                                            "desperately"))
    
    cleaned$sex<- cleaned$sex %>% gsub("m", "male", .) %>% gsub("f", "female",.)
    
    #ggplot(cleaned, aes(x = drinks, y = age)) + 
        #geom_violin(alpha = 0.3, size = 1, fill = "#FF9933", color = "darkred") + 
        #ggtitle("okcupid: drinks versus age") + facet_grid(sex~.)
    
    cleaned <- mutate(cleaned, age_range = 2 + 4 * age %/% 4 )
    
    drinks_and_sex <- group_by(cleaned, age_range, sex, drinks) %>% summarize(n=n())
    
    sex <- group_by(cleaned, age_range, sex) %>% summarize(NN=n())
    
    analyzed <- left_join(drinks_and_sex, sex)
    
    analyzed <- mutate(analyzed, percent = 100*n/NN)
    analyzed <- filter(analyzed, age_range >= 20)
    
    ## make color palette
    
    n.color <- length(unique(analyzed$drinks))
    getPalette = colorRampPalette(brewer.pal(8, "Dark2"))
    
    
    p <- ggplot(analyzed, aes(age_range, weight = percent)) +
        geom_bar(aes(fill = drinks), alpha = 0.9) +
        ggtitle("drinking and age") + facet_grid(sex~.) +
        coord_flip() + 
        scale_fill_manual(values = getPalette(n.color)) + 
        ylab("percent") + 
        xlab("age")
    
    print(p)
    
```

In this case we can get a better look at the data by using a semi-log plot. The strong decrease is heavy drinking is apparent for both sexes, though is faster for women than for men. There is an nteresting iuptick in "drinking often" for females in older age. 
    
```{r echo=4, fig.align='center', fig.height=5,fig.width=7.5, message = FALSE}   
    p3 <- ggplot(analyzed, aes(x = age_range, y = percent, color = drinks)) + geom_line(size = 1.2) +
        scale_color_manual(values = getPalette(n.color)) + 
        facet_grid(sex~.) +
        ggtitle("okcupid: drinking versus age") + 
        scale_y_log10(breaks = c(0.2, 0.5, 1, 2, 5, 10, 20, 50, 100))
    
    print(p3)
    
```  
##Sex and Religion

Surprisingly, men and women greatly differ in religious affiliation, with approximately 45% of men reprting to be either atheist or agnostic versus 35% for women.


```{r echo=FALSE, fig.align='center', fig.height=5,fig.width=7.5, message = FALSE}
 
    
    ## clean religion data into broad category
    bnG <- filter(profiles, !is.na(drinks), !is.na(religion), !is.na(sex)) %>% as_data_frame
    
    ## get affiliation (strip modifiers) using gsub and simple regex
    bnG$religious_affil <- gsub(' [A-z ]*', '', bnG$religion) %>% as.factor()
    
    ## group the data 
    bnG_rd <- group_by(bnG[,c("religious_affil", "sex")], religious_affil, sex) %>% summarize(n = n())
    
    bnG_r <- group_by(bnG[,c("religious_affil", "sex")], sex) %>% summarize(NN = n())
    
  
    bnG_final <- left_join(bnG_rd, bnG_r) %>% mutate(freq = n/NN, freq = ifelse(is.na(freq), 0, freq))
    
    
    
    # Source: local data frame [106 x 6]
    # Groups: drinks, religious_affil [54]
    # 
    #         drinks religious_affil   sex     n    NN        freq
    #         <fctr>          <fctr> <chr> <int> <dbl>       <dbl>
    # 1  desperately     agnosticism     f    25  3184 0.007851759
    # 2  desperately     agnosticism     m    37  5421 0.006825309
    # 3  desperately         atheism     f    18  1882 0.009564293
    # 4  desperately         atheism     m    40  4921 0.008128429
    # 5  desperately        buddhism     f     5   791 0.006321113
    # 6  desperately        buddhism     m    15  1090 0.013761468
    # 7  desperately     catholicism     f     4  2059 0.001942691
    # 8  desperately     catholicism     m     7  2613 0.002678913
    # 9  desperately    christianity     f     7  2606 0.002686109
    # 10 desperately    christianity     m     7  3052 0.002293578
    # # ... with 96 more rows
    

    ## express sex in actual words
    
    bnG_final$sex<- gsub("m", "male", bnG_final$sex)
    bnG_final$sex<- gsub("f", "female", bnG_final$sex)
    
    n.color <- length(unique(analyzed$sex))
    getPalette = colorRampPalette(brewer.pal(8, "Dark2"))
    
    p <- ggplot(bnG_final, aes(religious_affil, weight = 100*freq)) +
        geom_bar(aes(fill = sex), position = "dodge", alpha = 1) +
        ggtitle("okcupid: sex and religion") +
        scale_fill_manual(values = getPalette(n.color)) + 
        ylab("percent") + 
        xlab("religious affiliation") + 
        coord_flip()
    
    print(p)
    

```

##Income and Drinking

Surprisingly, men and women greatly differ in religious affiliation, with approximately 45% of men reprting to be either atheist or agnostic versus 35% for women.


```{r echo = FALSE, fig.align = 'center', fig.height = 3 , fig.width = 7.5 , message = FALSE}
 
    
    ## clean religion data into broad category
    cleaned <- filter(profiles, !is.na(income), !is.na(drinks), !is.na(sex)) %>% as_data_frame
    
    cleaned$income <- cleaned$income %>% as.factor()
    
    
    cleaned$drinks <- factor(cleaned$drinks, levels = c("not at all", 
                                                            "rarely",
                                                            "socially",
                                                            "often",
                                                            "very often",
                                                            "desperately"))
    
    ## group the data 
    income_and_drinks <- group_by(cleaned[,c("drinks", "sex", "income")], drinks, sex, income) %>% summarize(n = n())
    
    income <- group_by(cleaned[,c("drinks", "income", "sex")], income) %>% summarize(NN = n())
    
  
    analyzed <- left_join(income_and_drinks, income) %>% mutate(freq = n/NN, freq = ifelse(is.na(freq), 0, freq))
    
    

    ## express sex in actual words
    
    analyzed$sex<- analyzed$sex %>% gsub("m", "male", .) %>% gsub("f", "female", .)
    analyzed$drinks <- factor(analyzed$drinks, levels = c("not at all", 
                                                            "rarely",
                                                            "socially",
                                                            "often",
                                                            "very often",
                                                            "desperately"))
    
    n.color <- length(unique(analyzed$income))
    getPalette = colorRampPalette(brewer.pal(8, "Dark2"))
    
    p <- ggplot(analyzed, aes(income, weight = 100*freq)) +
        geom_bar(aes(fill = drinks), position = "dodge", alpha = 1) +
        ggtitle("drinking and income") +
        scale_fill_manual(values = getPalette(n.color)) + 
        ylab("percent") + 
        xlab("income") +
        coord_flip()
        
        #scale_y_log10(breaks = c(0.2, 0.5, 1, 2, 5, 10, 20, 50, 100))
    
    print(p)
    
```    

The most obvious in the graph above is that social drinkingm, the largest component of the rinking spectrum, shows an obvious arc, with the frequency of social drinkers peaking in the middle of the range and decreasing on the edges. We can get a better look at the 

```{r echo = FALSE, fig.align = 'center', fig.height = 3 , fig.width = 7.5 , message = FALSE}

    p3 <- ggplot(analyzed, aes(x = income, y = 100*freq, color = drinks, group = drinks)) + geom_line(size = 1.2) +
        scale_color_manual(values = getPalette(n.color)) + 
        facet_grid(sex~.) +
        ggtitle("drinking and income") + 
        scale_y_log10(breaks = c(0.2, 0.5, 1, 2, 5, 10, 20, 50, 100))
    
    print(p3)

```


Let's try looking at the sentiment of the texts

